# Data wrangling {-}


```{r, warning=F,message=F}
library(tidyverse)
library(tidyr)
library(ggthemes)
library(lubridate)

# Now that we have learned how to munge (manipulate) data
# and plot it, we will work on using these skills in new ways

# knitr::opts_knit$set(root.dir='..')
```


```{r dataread, warning=F,message=F}
####-----Reading in Data and Stacking it ----- ####
#Reading in files
files <- list.files('dataDataWrangle',full.names=T)

```

```{r}
#Read in individual data files
ndmi <- read_csv(files[1]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')

ndsi <- read_csv(files[2]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')

ndvi <- read_csv(files[3])%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')

# Stack as a tidy dataset
full_long <- rbind(ndvi,ndmi,ndsi) %>%
  gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value))


```



## Question 1


1) What is the correlation between NDVI and NDMI? 
Here, I want you to convert the full_long dataset in to a wide dataset using the function "spread."
Then, make a plot that shows the correlation as a function of if the site was burned or not (x axis should be ndmi).
You should exclude winter months and focus on summer months.


```{r longformat}
# Convert from long to wide data with spread() 
# Add month and year columns to wide data
full_wide1 <- full_long %>%
  spread(key='data', value='value') %>% 
  mutate(month=month(DateTime)) %>% 
  mutate(year=year(DateTime))
```


```{r plot ndvi vs ndmi, warning=F,message=F}
# Plot ndvi as response and ndmi as predictor, distinguishing between burn or unburned sites, limit to only summer months 
summer_wide <- full_wide1 %>%
  filter(month %in% c(6,7,8))
ggplot(summer_wide, aes(x=ndmi, y=ndvi, color=site))+
  geom_point(shape=1)+
  labs(x="NDMI", y="NDVI")+
  theme_few()+ 
  scale_color_manual(name="Site",labels=c("Burned","Unburned"), values= c("#274a12","#babf28"))+ 
  xlim(-0.6,0.7)+ 
  ylim(0.05, 0.6)
```


```{r plot ndvi vs ndmi, facet wrap site}
# Plot ndvi vs ndmi with facet wrap over site type (burned or unburned)
ggplot(summer_wide, aes(x=ndmi, y=ndvi))+
  geom_point(shape=1)+
  labs(x="NDMI", y="NDVI")+
  theme_few()+
  facet_wrap('site')
```


```{r}
# Test correlation between ndvi and ndmi
cor.test(summer_wide$ndmi, summer_wide$ndvi, method='pearson')
# Fit lm model for ndvi by ndmi
LMFit1 <- lm(ndvi~ndmi, data =summer_wide)
summary(LMFit1)
```


```{r LMFit1 diagnostics, echo = TRUE, fig.show='hide'}
# LMFit1 diagnostic plots 
par(mfrow=c(1,2))
plot(LMFit1, which= c(1:2))
```



**1. Based upon a test of correlation, we have evidence of a positive linear association between summer NDMI and summer NDVI, with a p-value < 2.2e-16 (less than 0.05). For every 1 unit increase in summer NDMI, there is a 0.908772 increase in summer NDVI (p-value < 2.2e-16).** 



#### Bonus Question 1
```{r convert long to wide via pivot_wider()}
# Convert from long to wide with pivot_wider() 
full_wide2 <- full_long %>%
  pivot_wider(names_from = 'data', values_from = 'value')
```


## Question 2 


2) What is the correlation between average NDSI (normalized snow index) for January - April and average NDVI for June-August?
In other words, does the previous year's snow cover influence vegetation growth for the following summer?


```{r, warning=F,message=F}
# Summarize data by average ndvi for summer months
summer_ndvi<- full_wide1 %>%
  group_by(site, year, month) %>%
  filter(month %in% c(6, 7, 8)) %>%
  summarize(mean_ndvi = mean(ndvi)) %>%
  filter(!is.na(mean_ndvi))

# Summarize data by average ndsi over winter months 
winter_ndsi <- full_wide1 %>%
  group_by(site, year, month) %>%
  filter(month %in% c(1, 2, 3, 4)) %>%
  summarize(mean_ndsi = mean(ndsi)) %>%
  filter(!is.na(mean_ndsi))
```


```{r}
# Join average summer ndvi and average winter ndsi by year and site
# Add burnperiod column to distinguish pre- and post-fire years
wide_averages <- inner_join(winter_ndsi, summer_ndvi, by= c('site', 'year')) %>%
  mutate(burnperiod = as.factor(ifelse(year < 2002,"prefire", "postfire"))) 
```

  
```{r}
# Plot and evaluate relationship between ndvi and ndsi
ggplot(wide_averages, aes(x=mean_ndsi, y=mean_ndvi))+
  geom_point(shape=1, color = "#22AA99")+
  geom_smooth(method=lm, color="#22AA99", size=0.1, se=FALSE)+
  theme_few()+ 
  labs(x="Average Winter NDSI", y="Average Summer NDVI")
```


```{r}
# Test correlation between ndvi and ndsi overall
cor.test(wide_averages$mean_ndsi, wide_averages$mean_ndvi)
# Fit lm model for ndvi by ndsi overall
LMFit2 <- lm(mean_ndvi~mean_ndsi, data = wide_averages)
summary(LMFit2)
```


```{r LMFit2 diagnostics, echo = TRUE, fig.show='hide'}
# LMFit2 diagnostics
par(mfrow=c(1,2))
plot(LMFit2, which= c(1:2))
```



**2. The p-value from a test of correlation for average summer NDVI and average winter NDSI is 0.0002124 and less than 0.05. We have evidence of a positive linear association between average summer NDVI and average winter NDSI. For every 1 unit increase in average winter NDSI, there is a 0.042658 increase in average summer NDVI.** 



## Question 3


3) How is the snow effect from question 2 different between pre- and post-burn and burned and unburned? 


#### *NDVI-NDSI: Pre- and postfire figure and analyses* 

```{r}
# Plot and compare ndvi-ndsi relationship between pre- and post-burn periods
ggplot(wide_averages, aes(x=mean_ndsi, y=mean_ndvi, color=burnperiod))+
  geom_point(shape=1)+
  theme_few()+
  geom_smooth(method=lm, size=0.1, se=FALSE)+
  labs(x="Average Winter NDSI", y="Average Summer NDVI")+
  scale_color_manual(name="",labels=c("Pre-fire","Post-fire"), values= c("#2f94b5","#b5982f"))
```


```{r, echo = TRUE, results='hide'}
# Fit lm model to evaluate ndvi-ndsi correlation pre- and post-fire 
LMFit3 <- lm(mean_ndvi~mean_ndsi*burnperiod, data= wide_averages)
summary(LMFit3)
```


```{r LMFit3 diagnostics, echo = TRUE, fig.show='hide'}
# LMFit3 diagnostic plots
par(mfrow=c(1,2))
plot(LMFit3, which= c(1:2))
```


```{r}
# Create separate data frames for prefire and postfire 
prefire <- wide_averages %>% 
  filter(burnperiod %in% 'prefire')
  
postfire <- wide_averages %>% 
  filter(burnperiod %in% 'postfire')
```


```{r}
# Test correlation between ndvi and ndsi prefire and postfire 
cor.test(prefire$mean_ndsi,prefire$mean_ndvi)
cor.test(postfire$mean_ndsi,postfire$mean_ndvi)
```



**3a. In pre-fire years, we do not have evidence of an association between average summer NDVI and average winter NDSI (p-value=0.1721). However, in post-fire years, we have evidence of an association between average summer NDVI and average winter NDSI (p-value=0.01048).** 



#### *NDVI-NDSI: Burned and unburned figure and analyses*

```{r}
# Plot and compare ndvi-ndsi relationship across burned versus unburned sites
ggplot(wide_averages, aes(x=mean_ndsi, y=mean_ndvi, color=site))+
  geom_point(shape=1)+
  theme_few()+ 
  geom_smooth(method=lm, size=0.1, se=FALSE)+
  labs(x="Average Winter NDSI", y="Average Summer NDVI")+ 
  scale_color_manual(name="Site",labels=c("Burned","Unburned"), values= c("#292423","#a4a823"))
```


```{r, echo = TRUE, results='hide'}
# Fit lm model to evaluate ndvi-ndsi correlation between burned and unburned sites
LMFit4 <- lm(mean_ndvi~mean_ndsi*site, data=wide_averages)
summary(LMFit4)
```


```{r LMFit4 diagnostics, echo = TRUE, fig.show='hide'}
# LMFit4 diagnostics
par(mfrow=c(1,2))
plot(LMFit4, which= c(1:2))
```


```{r}
# Create separate data frames for burned and unburned sites
burned <- wide_averages %>% 
  # filter(burnperiod %in% 'postfire') %>%
  filter(site %in% 'burned')
  
unburned <- wide_averages %>% 
  # filter(burnperiod %in% 'postfire') %>%
  filter(site %in% 'unburned')
```


```{r}
# Test correlation between ndvi and ndsi in burned and unburned sites separately 
cor.test(burned$mean_ndsi, burned$mean_ndvi)
cor.test(unburned$mean_ndsi,unburned$mean_ndvi)
```



**3b. When analyzing all years included in the dataset, we do not have evidence of an association between average summer NDVI and average winter NDSI within the unburned area (p-value= 0.6589) and burned area (p-value=0.1825). When analyzing only years postfire, we still do not have evidence of an association between average summer NDVI and average winter NDSI in the unburned (p-value=0.905) and burned areas (p-value=0.3226).**



## Question 4


4) What month is the greenest month on average? 


```{r}
# Calculate maximum, monthly mean ndvi over time
ndvi_stats<- wide_averages %>% 
  group_by(month.y) %>% 
  summarize(max_ndvi=max(mean_ndvi))
ndvi_stats
```



**4. August is the greenest month on average.** 



## Question 5 


5) What month is the snowiest on average?


```{r}
# Calculate maximum, monthly mean ndsi over time
ndsi_stats <- wide_averages %>% 
  group_by(month.x) %>% 
  summarize(max_ndsi=max(mean_ndsi))
ndsi_stats
```



**5. February is the snowiest month on average.** 



#### Bonus Question 2: Use Climate Engine to pull the same data for the assignment, but updated with 2020/2021 data.

Updated Climate Engine with years 2020 and 2021 is stored in data2. 


