# Key programming concepts  


```{r, include=FALSE}
#Load libraries
library(rvest)
library(tidyverse)
library(lubridate)
library(readxl)
library(ggthemes)
library(tibble)
library(pdftools)
library(dygraphs)
library(xts)
library(lubridate)
library(kableExtra)
```


## Introduction


This project served as an introduction to webscraping, iteration, and  functions. We extracted data from the Center for Snow and Avalanche Studies (CSAS) [website](https://snowstudies.org/archived-data/): home to incredibly rich snow, temperature, and precipitation data. 

Project objectives: 

a) Extract data URLS or CSV links from website
b) Download data using a for loop or map function 
c) Write custom functions


## Data Acquisiton: Snowpack Depth 


#### Webscraping: extract links from webpage {-}


```{r}
#Save site URL
site_url <- 'https://snowstudies.org/archived-data/'

#Read the web URL
webpage <- read_html(site_url)

#Extract only weblinks and then URLs
links <- webpage %>%
  html_nodes('a') %>%
  .[grepl('24hr',.)] %>%
  html_attr('href')
```


```{r, include=F}
#Code for extracting from tables (didn't work)
tables <- webpage %>%
  html_nodes('table') %>%
  magrittr::extract2(3) %>%
  html_table(fill = TRUE)
```


#### Download data using a for loop {-}


```{r, cache=TRUE}
#Split by forward slashes to parse out name of file
splits <- str_split_fixed(links,'/',8)

#Select column containing file name
dataset <- splits[,8] 

#Generate a file list to hold the data
file_names <- paste0('dataFunctions/',dataset)

#Download data in a for loop
for(i in 1:3){
  download.file(links[i],destfile=file_names[i])
}

downloaded <- file.exists(file_names)

evaluate <- !all(downloaded)
```


#### Download data using map function(s) {-}


```{r, cache=T, results='hide'}
#Utilize map for same operation as above, download the same 3 files 
if(evaluate == T) {
  map2(links[1:3], file_names[1:3], download.file)
} else{
  print('data already downloaded')
}
```


#### Read in snow data with pattern matching {-}


```{r}
#Read in only snow data as a loop
#Use pattern matching to only keep certain files
snow_files <- file_names %>%
  .[!grepl('SG_24',.)] %>%
  .[!grepl('PTSP',.)]
```


```{r, include=F}
#empty_data <- list()

# snow_data <- for(i in 1:length(snow_files)){
#   empty_data[[i]] <- read_csv(snow_files[i]) %>%
#     select(Year,DOY,Sno_Height_M)
# }

#snow_data_full <- do.call('rbind',empty_data)

#summary(snow_data_full)
```


#### Read in snow data with map function {-}


```{r, results='hide', message=F}
#Create function to read in snow files
our_snow_reader <- function(file){
  name = str_split_fixed(file,'/',2)[,2] %>%
    gsub('_24hr.csv','',.)
  df <- read_csv(file) %>%
    select(Year,DOY,Sno_Height_M) %>%
    mutate(site = name)
}

snow_data_full <- map_dfr(snow_files,our_snow_reader)
```


## Results: Snowpack Depth 


```{r, message=F, warning=FALSE}
#Create column for yearly mean snow height by site
snow_yearly <- snow_data_full %>%
  group_by(Year,site) %>%
  summarize(mean_height = mean(Sno_Height_M,na.rm=T))

#Plot yearly mean snow height by site
ggplot(snow_yearly,aes(x=Year,y=mean_height,color=site)) + 
  geom_point(alpha=0.75) +
  ggthemes::theme_few() + 
  scale_color_manual(name="Study Plots", labels=c("Swamp Angel","Senator Beck"), values= c("#58A3EB","#58E8EB"))+
  labs(y="Depth of Snowpack (m)",main="Average Snow Height over Time")
```

<small> **Figure 1. Yearly snow pack depth for Swamp Angel and Senator Beck study plots, Center for Snow and Avalanche Studies. Height of snow was sampled once at the end of an array.** </small>


## Data Acquisition: Meterological Data 


### Webscraping: extract meterological data URLs {-}


```{r}
#Extract data URLs for SASP forcing and SBSP forcing datasets
site_url2 <- 'https://snowstudies.org/archived-data/'

webpage2 <- read_html(site_url2)

links2 <- webpage2 %>% 
  html_nodes('a') %>% 
  .[grepl('forcing',.)] %>% 
  html_attr('href')
```


### Download meteorological data {-}


```{r, results='hide'}
#Download the meterological data and save within data folder 
splits2 <- str_split_fixed(links2,'/',8)

dataset2 <- splits2[,8]

filenames2 <- paste0('dataFunctions/',dataset2)

#Utilize map2() to download data 
map2(links2,filenames2,download.file)
```


```{r include=F}
#Download data in a for loop 
# for (i in 1:length(filenames2)) {
#   download.file(links2[i], destfile = filenames2[i])
# }
```


### Write a custom function to read in data {-}


```{r}
#Retrieve variable names from the metadata pdf file
headers <- pdf_text('https://snowstudies.org/wp-content/uploads/2022/02/Serially-Complete-Metadata-text08.pdf') %>%
  readr::read_lines(.) %>%
  trimws(.) %>%
  str_split_fixed(.,'\\.',2) %>%
  .[,2] %>%
  .[1:26] %>%
  str_trim(side = "left")

#Write a function to read in the data and append site column 
forcing_files <- filenames2  

file <- forcing_files[1]

forcefile_reader <- function(file) {
  name2 = str_split_fixed(file, '_', 3)[, 2] 
    df <- read.csv(file, header = FALSE, sep = '') %>%
      select(V1, V2, V3, V7, V10) %>%
      rename(
        year = 1,
        month = 2,
        day = 3,
        precip = 4,
        airtemp = 5
      ) %>%
      mutate(site = name2)
    
}
```


### Read in meterological data files with map function {-}


```{r}
#Use map function to read in meteorological files 
forcing_data_full <- map_dfr(forcing_files, forcefile_reader)

#Display summary as tibble 
forcing_tibble <-as_tibble(forcing_data_full)
kable(forcing_tibble)
```


## Results: Meterological Data 


```{r, message=F}
#Create data frame with mean air temperature by year by site 
temp_yearly <- forcing_data_full %>% 
  filter(!year %in% 2003) %>% 
  group_by(year,site) %>% 
  summarize(mean_yrtemp = mean(airtemp, na.rm=T))
  
#Make line plot of mean temperature by year by site 
ggplot(temp_yearly, aes(x=year, y=mean_yrtemp, color=site))+
  geom_line(size=.75)+ 
  theme_few()+ 
  scale_color_manual(name="Study Plots", labels=c("Swamp Angel","Senator Beck"), values= c("#58A3EB","#58E8EB"))+ 
  labs(x='Year', y='Average Air Temperature (K)')
```


<small> **Figure 2. Yearly mean air temperature for Swamp Angel and Senator Beck study plots, CSAS. Records from 2003 were excluded given data was only available for two months out of the year.** </small>


```{r, message=F}
#Write function to make line plots of monthly average temps per site per year
line_plotter <- function(df, year) {
  temp_monthly <- df %>%
    group_by(year, month, site) %>%
    summarize(mean_motemp = mean(airtemp, na.rm = T)) %>%
    filter(i == year)
  
  print(
    ggplot(temp_monthly, aes(
      x = month, y = mean_motemp, color = site
    )) +
      geom_line(size = .75) +
      theme_few() +
      scale_color_manual(name="Study Plots", labels=c("Swamp Angel","Senator Beck"), values = c("#762448", "#B1D374")) +
      labs(x="Month",y = 'Average Air Temperature (K)', title = i)+
      scale_x_continuous(
        breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
        labels = c('January','February','March','April','May','June','July','August','September','October','November','December')) +
      theme(
        axis.text.x = element_text(
          color = "black",
          size = 8,
          angle = 30,
          vjust = .8,
          hjust = .8
        )
      ))
}

#Use for loop to plot years 2005 to 2010
yrs = c(2005:2010)

for (i in yrs){
  line_plotter(forcing_data_full,year)
}
```


<small> **Figures 3-9. Monthly average temperature at study plots: Swamp Angel and Senator Beck, years 2005-2010.** </small>


```{r, message=F}
#Add date column to data frame
#Use lubridate:: yday for day of year 
precip_daily <- forcing_data_full %>%
  group_by(month, day, year, site) %>%
  summarize(mean_precip = mean(precip, na.rm = T)) %>%
  mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
  mutate(yday = yday(date)) %>%
  pivot_wider(names_from = site, values_from = mean_precip) %>%
  dplyr::select(-SBSP)%>% 
  group_by(yday)%>% 
  summarize(mean_dy_precip = mean(SASP))

#Create plot with mean daily precip by day of year using ggplot 
ggplot(precip_daily, aes(x = yday, y = mean_dy_precip)) +
  geom_point(alpha=0.5) +
  theme_few() +
  labs(x = 'Day of Year', y = 'Average Precipitation (mm/day)')+
  scale_x_continuous(breaks=c(1,90,180,270,360))
```


<small> **Figure 10. Average daily precipitation by day of year at the Swamp Angel study plot, averaged across available years.** </small>


```{r, message=F}
#Create plot with mean daily precip by day of year using dygraphs
#This is not averaged across available years 

precip_daily2 <- forcing_data_full %>%
  group_by(month, day, year, site) %>%
  summarize(mean_precip = mean(precip, na.rm = T)) %>%
  mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
  pivot_wider(names_from = site, values_from = mean_precip) %>%
  ungroup() %>%
  select(-SBSP, -month, -day, -year)

precip_xts2 <- xts(precip_daily2 %>%
                     select(SASP), order.by = precip_daily2$date)

dygraph(precip_xts2, ylab = "") %>%
  dyOptions(fillGraph = TRUE, axisLabelFontSize=10) %>% 
  dySeries("SASP", label = "Average Daily Precipitation")
```


<small> **Figure 11. Interactive plot of average precipitation by day of year at Swamp Angel study plot.** </small>


```{r ggplot, yearly precip plot, message=F}
#Write a function to create yearly plots for precip by day of year with ggplot
precip_plotter2 <- function(df, year) {
  precip_daily4 <- df %>%
    group_by(month, day, year, site) %>%
    summarize(mean_precip = mean(precip, na.rm = T)) %>%
    mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
    mutate(yday = yday(date)) %>%
    pivot_wider(names_from = site, values_from = mean_precip) %>%
    select(-SBSP) %>%
    filter(year == i)
  
  print(
    ggplot(precip_daily4, aes(x = yday, y = SASP)) +
      geom_point(shape = 1) +
      theme_few() +
      labs(
        title = i,
        x = 'Day of Year',
        y = 'Average Precipitation (mm/day)') +
      scale_x_continuous(breaks = c(1, 90, 180, 270, 360))
  )
}

#Use for loop to plot years 2005 to 2010
yrs = c(2005:2010)

for (i in yrs){
  precip_plotter2(forcing_data_full,year)
}
```


<small> **Figures 12-17. Yearly plots (2005-2010) of precipitation by day of year at the Swamp Angel study plot.** </small> 


```{r dygraphs, yearly precip plots, include=F}
# #Write function to create yearly plots of precipitation by day of year with dygraphs
# dy_precip_plotter <- function(df, year) {
#   precip_daily3 <- df %>%
#     group_by(month, day, year, site) %>%
#     summarize(mean_precip = mean(precip, na.rm = T)) %>%
#     mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
#     pivot_wider(names_from = site, values_from = mean_precip) %>%
#     filter(year == i) %>%
#     ungroup() %>%
#     select(-SBSP,-month,-day,-year)
#   
#   precip_xts3 <- xts(precip_daily3 %>%
#                       select(SASP), order.by = precip_daily3$date)
#   
#   print(
#     dygraph(precip_xts3, main = i, ylab = "Average Daily Precipitation") %>%
#       dyOptions(fillGraph = TRUE, axisLabelFontSize=10)
#   )
# }
# 
# #Use for loop to plot years 2005 to 2010
# yrs = c(2005:2010)
#  
# for (i in yrs){
#   dy_precip_plotter(forcing_data_full,year)
# }
```

