# Key programming concepts: webscraping, iteration, functions {-}


```{r, include=FALSE}
library(rvest)
library(tidyverse)
library(lubridate)
library(readxl)

```


# Simple web scraping

R can read html using either rvest, xml, or xml2 packages. Here we are going to navigate to the Center for Snow and Avalance Studies  [Website](https://snowstudies.org/archived-data/) and read a table in. This table contains links to data we want to programmatically download for three sites. We don't know much about these sites, but they contain incredibly rich snow, temperature, and precip data. 


## Reading an html 

### Extract CSV links from webpage

```{r}
site_url <- 'https://snowstudies.org/archived-data/'

#Read the web url
webpage <- read_html(site_url)

#See if we can extract tables and get the data that way
tables <- webpage %>%
  html_nodes('table') %>%
  magrittr::extract2(3) %>%
  html_table(fill = TRUE)
#That didn't work, so let's try a different approach

#Extract only weblinks and then the URLs!
links <- webpage %>%
  html_nodes('a') %>%
  .[grepl('24hr',.)] %>%
  html_attr('href')

```


## Data Download

### Download data in a for loop

```{r}

#Grab only the name of the file by splitting out on forward slashes
splits <- str_split_fixed(links,'/',8)

#Keep only the 8th column
dataset <- splits[,8] 

#generate a file list for where the data goes
file_names <- paste0('dataFunctions/',dataset)

for(i in 1:3){
  download.file(links[i],destfile=file_names[i])
}

downloaded <- file.exists(file_names)

evaluate <- !all(downloaded)

```


### Download data in a map

```{r}

#Map version of the same for loop (downloading 3 files)
if(evaluate == T){
  map2(links[1:3],file_names[1:3],download.file)
}else{print('data already downloaded')}

```


## Data read-in 

### Read in just the snow data as a loop

```{r}
#Pattern matching to only keep certain files
snow_files <- file_names %>%
  .[!grepl('SG_24',.)] %>%
  .[!grepl('PTSP',.)]

#empty_data <- list()

# snow_data <- for(i in 1:length(snow_files)){
#   empty_data[[i]] <- read_csv(snow_files[i]) %>%
#     select(Year,DOY,Sno_Height_M)
# }

#snow_data_full <- do.call('rbind',empty_data)

#summary(snow_data_full)
```


### Read in the data as a map function

```{r}

our_snow_reader <- function(file){
  name = str_split_fixed(file,'/',2)[,2] %>%
    gsub('_24hr.csv','',.)
  df <- read_csv(file) %>%
    select(Year,DOY,Sno_Height_M) %>%
    mutate(site = name)
}

snow_data_full <- map_dfr(snow_files,our_snow_reader)

summary(snow_data_full)
```


### Plot snow data

```{r}
snow_yearly <- snow_data_full %>%
  group_by(Year,site) %>%
  summarize(mean_height = mean(Sno_Height_M,na.rm=T))

ggplot(snow_yearly,aes(x=Year,y=mean_height,color=site)) + 
  geom_point() +
  ggthemes::theme_few() + 
  ggthemes::scale_color_few()
```



# Assignment:


1. Extract the meteorological data URLs. Here we want you to use the `rvest` package to get the URLs for the `SASP forcing` and `SBSP_forcing` meteorological datasets.


### Extract meteorological data URLS from website

```{r}
# Extract data URLs for SASP forcing and SBSP forcing datasets

site_url2 <- 'https://snowstudies.org/archived-data/'

webpage2 <- read_html(site_url2)

links2 <- webpage2 %>% 
  html_nodes('a') %>% 
  .[grepl('forcing',.)] %>% 
  html_attr('href')

links2
```



2. Download the meteorological data. Use the `download_file` and `str_split_fixed` commands to download the data and save it in your data folder. You can use a for loop or a map function. 


### Download data and save it within data folder 

```{r}
# Download the data and save within data folder 

splits2 <- str_split_fixed(links2,'/',8)

dataset2 <- splits2[,8]

filenames2 <- paste0('dataFunctions/',dataset2)

# Download data in a for loop 

# for (i in 1:length(filenames2)) {
#   download.file(links2[i], destfile = filenames2[i])
# }

# Utilize map2() to download data 

map2(links2,filenames2,download.file)

```



3. Write a custom function to read in the data and append a site column to the data. 


### Write function to read in data and append site column to the data

```{r}

# This code grabs the variable names from the metadata pdf file

library(pdftools)
headers <- pdf_text('https://snowstudies.org/wp-content/uploads/2022/02/Serially-Complete-Metadata-text08.pdf') %>%
  readr::read_lines(.) %>%
  trimws(.) %>%
  str_split_fixed(.,'\\.',2) %>%
  .[,2] %>%
  .[1:26] %>%
  str_trim(side = "left")
headers

# Write a function to read in the data and append site column 

forcing_files <- filenames2  
#   .[!grepl('24hr',.)]
# forcing_files


file <- forcing_files[1]


forcefile_reader <- function(file) {
  name2 = str_split_fixed(file, '_', 3)[, 2] 
    df <- read.csv(file, header = FALSE, sep = '') %>%
      select(V1, V2, V3, V7, V10) %>%
      rename(
        year = 1,
        month = 2,
        day = 3,
        precip = 4,
        airtemp = 5
      ) %>%
      mutate(site = name2)
    
}

```



4. Use the `map` function to read in both meteorological files. Display a summary of your tibble.


### Read in meterological data files with map function 

```{r}

# Use map function to read in meteorological files 

forcing_data_full <- map_dfr(forcing_files, forcefile_reader)

# # Display summary as tibble 

library(tibble)
forcing_tibble <-as_tibble(forcing_data_full)
summary(forcing_tibble)

```



5. Make a line plot of mean temp by year by site (using the `air temp [K]` variable). Is there anything suspicious in the plot? Adjust your filtering if needed.


### Make line plot of mean temperature by year by site 

```{r}
library(ggthemes)

# Create data frame with mean air temperature by year by site 

temp_yearly <- forcing_data_full %>% 
  filter(!year %in% 2003) %>% 
  group_by(year,site) %>% 
  summarize(mean_yrtemp = mean(airtemp, na.rm=T))
  
# Make line plot of mean temperature by year by site 

ggplot(temp_yearly, aes(x=year, y=mean_yrtemp, color=site))+
  geom_line(size=.75)+ 
  theme_few()+ 
  scale_color_manual(name="Site", values= c("#58A3EB","#58E8EB"))+ 
  labs(x='Year', y='Average Air Temperature (K)')
```


I filtered out 2003 given there were only records for two months out of the year. 


6. Write a function that makes line plots of monthly average temperature at each site for a given year. Use a for loop to make these plots for 2005 to 2010. Are monthly average temperatures at the Senator Beck Study Plot ever warmer than the Snow Angel Study Plot?
Hint: https://ggplot2.tidyverse.org/reference/print.ggplot.html


### Create function to make line plots of monthly average temperature at each site per year

```{r}

# Write function to make line plots of monthly average temps per site per year

line_plotter <- function(df, year) {
  temp_monthly <- df %>%
    group_by(year, month, site) %>%
    summarize(mean_motemp = mean(airtemp, na.rm = T)) %>%
    filter(i == year)
  
  print(
    ggplot(temp_monthly, aes(
      x = month, y = mean_motemp, color = site
    )) +
      geom_line(size = .75) +
      theme_few() +
      scale_color_manual(name = "Site", values = c("#762448", "#B1D374")) +
      labs(x="Month",y = 'Average Air Temperature (K)', title = i)+
      scale_x_continuous(
        breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
        labels = c(
          'January',
          'February',
          'March',
          'April',
          'May',
          'June',
          'July',
          'August',
          'September',
          'October',
          'November',
          'December'
        )
      ) +
      theme(
        axis.text.x = element_text(
          color = "black",
          size = 8,
          angle = 30,
          vjust = .8,
          hjust = .8
        )
      ))
}

# Use for loop to plot years 2005 to 2010

yrs = c(2005:2010)

for (i in yrs){
  line_plotter(forcing_data_full,year)
}

```


Bonus: Make a plot of average daily precipitation by day of year (averaged across all available years).  


### Create a plot of average daily precip by day of year

```{r}

# Create plot with mean daily precip by day of year using ggplot 

# Add date column 
# Use lubridate:: y day for day of year 
precip_daily <- forcing_data_full %>%
  group_by(month, day, year, site) %>%
  summarize(mean_precip = mean(precip, na.rm = T)) %>%
  mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
  mutate(yday = yday(date)) %>%
  pivot_wider(names_from = site, values_from = mean_precip) %>%
  dplyr::select(-SBSP)%>% 
  group_by(yday)%>% 
  summarize(mean_dy_precip = mean(SASP))

ggplot(precip_daily, aes(x = yday, y = mean_dy_precip)) +
  geom_point(shape = 1) +
  theme_few() +
  labs(x = 'Day of Year', y = expression('Average Precipitation'~ ('kg'*m^2*s^1)))+
  scale_x_continuous(breaks=c(1,90,180,270,360))

# Create plot with mean daily precip by day of year using dygraphs
# This is not averaged across available years 

library(dygraphs)
library(xts)
library(lubridate)

precip_daily2 <- forcing_data_full %>%
  group_by(month, day, year, site) %>%
  summarize(mean_precip = mean(precip, na.rm = T)) %>%
  mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
  pivot_wider(names_from = site, values_from = mean_precip) %>%
  ungroup() %>%
  select(-SBSP, -month, -day, -year)

precip_xts2 <- xts(precip_daily2 %>%
                     select(SASP), order.by = precip_daily2$date)

dygraph(precip_xts2, ylab = "Average Daily Precipitation") %>%
  dyOptions(fillGraph = TRUE, axisLabelFontSize=10)

```



Bonus #2: Use a function and for loop to create yearly plots of precipitation by day of year.


### Create yearly plots of precip by day of year 

#### Function using ggplot 

```{r ggplot, yearly precip plot}

# Write a function to create yearly plots for precip by day of year with ggplot

precip_plotter2 <- function(df, year) {
  precip_daily4 <- df %>%
    group_by(month, day, year, site) %>%
    summarize(mean_precip = mean(precip, na.rm = T)) %>%
    mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
    mutate(yday = yday(date)) %>%
    pivot_wider(names_from = site, values_from = mean_precip) %>%
    select(-SBSP) %>%
    filter(year == i)
  
  print(
    ggplot(precip_daily4, aes(x = yday, y = SASP)) +
      geom_point(shape = 1) +
      theme_few() +
      labs(
        title = i,
        x = 'Day of Year',
        y = expression('Average Precipitation' ~ ('kg' *
                                                    m ^ 2 * s ^ 1))
      ) +
      scale_x_continuous(breaks = c(1, 90, 180, 270, 360))
  )
}

# Use for loop to plot years 2005 to 2010

yrs = c(2005:2010)

for (i in yrs){
  precip_plotter2(forcing_data_full,year)
}

```


#### Function using dygraphs

This will print the plots in the console/inline and the document will knit. However, the dygraphs plots do not show up in the final document. 


```{r dygraphs, yearly precip plots}

# Write function to create yearly plots of precipitation by day of year with dygraphs

dy_precip_plotter <- function(df, year) {
  precip_daily3 <- df %>%
    group_by(month, day, year, site) %>%
    summarize(mean_precip = mean(precip, na.rm = T)) %>%
    mutate(date = as.Date(paste(year, month, day, sep = "/"))) %>%
    pivot_wider(names_from = site, values_from = mean_precip) %>%
    filter(year == i) %>%
    ungroup() %>%
    select(-SBSP,-month,-day,-year)
  
  precip_xts3 <- xts(precip_daily3 %>%
                      select(SASP), order.by = precip_daily3$date)
  
  print(
    dygraph(precip_xts3, main = i, ylab = "Average Daily Precipitation") %>%
      dyOptions(fillGraph = TRUE, axisLabelFontSize=10)
  )
}

# Use for loop to plot years 2005 to 2010

yrs = c(2005:2010)
 
for (i in yrs){
  dy_precip_plotter(forcing_data_full,year)
}
```

