# Geospatial Analysis 


```{r, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(RApiSerialize)
library(ggthemes)
library(scales)
library(leaflet)
```


## Introduction


This project served as an introduction to spatial analysis in R and consisted of two parts. In part I, we utilized data from the LAGOS program to investigate the spatial distribution and size of lakes in three midwestern US states: Illinois, Iowa, and Minnesota. 

Project objectives: 

a) Use mapview package to create interactive maps 

b) Utilize sf package to work with geospatial vector data

c) Assess the distribution of lake size across states 


## Methods 


### Data Acquisition


```{r data-read}
#Download LAGOS data
# LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())

#Load in LAGOS data
lagos <- lagosne_load()

#Parse out lake centroid information
lake_centers <- lagos$locus
# load('lake_centers.Rdata')
```


### Data Processing


```{r include=F}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))
```


```{r}
#Create spatial object for lakes 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

#Subset number of lakes for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)
```


```{r}
#Create states variable 
states <- us_states()

#Select only Minnesota from states data 
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]
```


```{r, include=F}
#Plot all the states to check if they loaded
#mapview(states)
```


## Results


```{r}
#Plot the first 1000 Minnesota lakes
#Color by lake area in hectares
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha', at=c(0,500,1500,3000,7500,15000,30000,65000,125000),layer.name='Lake area (ha)')+ mapview(minnesota,layer.name="Minnesota", alpha.regions=0.35, col.regions="#fcf87f")
```

<small> Figure 1. Interactive map of Minnesota Lakes, colored by lake area in hectares. </small> 


```{r, include=F}
#Show a map outline of Iowa and Illinois 
i_states <- states %>% 
  filter(name %in% c('Iowa','Illinois')) %>% 
  st_transform(2163)
```


```{r}
#Create interactive plot of Iowa and Illinois lakes 
#Color by lake area in hectares
istate_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) %>% 
  mapview(., zcol = 'lake_area_ha', at=c(0,100,250,500,1000,2500,5000,10000), layer.name='Lake area (ha)')+ mapview(i_states,zcol='name', alpha.regions=0.35, col.regions=c("#DAA520","#9F2B68"), layer.name=F)
```


<small> Figure 2. Interactive map of Iowa and Illinois lakes, colored by lake area in hectares. </small> 


```{r, include=F}
#Subset LAGOS data
istate_lakes <- spatial_lakes[i_states,]

nrow(minnesota_lakes)-nrow(istate_lakes)
```


```{r, include=F}
#Create histogram of Minnesota lake size
ggplot(minnesota_lakes, aes(x=lake_area_ha))+ 
  geom_histogram(bins=40, alpha=0.25, color="black", fill="#03D1E2")+ 
  scale_x_log10(labels = label_comma(accuracy = 1))+
  theme_few()+
  labs(x= "Lake size (hectares)", y= "Frequency", title='Minnesota Lakes')
```


<small> Figure 3. Distribution of Minnesota lake size. </small> 


```{r, include=F}
# states_lagos <- lagos$state 
iowa <- states %>%
  filter(name == 'Iowa') %>%
  st_transform(2163)

#Subset lakes based upon spatial position
iowa_lakes <- spatial_lakes[iowa,]
  
#Create histogram of Iowa lake size 
ggplot(iowa_lakes, aes(x=lake_area_ha))+ 
  geom_histogram(bins=40, alpha=0.25, color="black", fill="#0368E2")+ 
  theme_few()+
  scale_x_log10(labels = label_comma(accuracy = 1))+
  labs(x= "Lake size (hectares)", y= "Frequency", title='Iowa Lakes')
```


<small> Figure 4. Distribution of Iowa lake size in hectares. </small>


```{r, include=F}
illinois <- states %>%
  filter(name == 'Illinois') %>%
  st_transform(2163)

#Subset lakes based upon spatial position
illinois_lakes <- spatial_lakes[illinois,]
  
#Create histogram of Iowa lake size 
ggplot(illinois_lakes, aes(x=lake_area_ha))+ 
  geom_histogram(bins=40, alpha=0.25, color="black", fill="#5D3FD3")+ 
  theme_few()+
  scale_x_log10(labels = label_comma(accuracy = 1))+
  labs(x= "Lake size (hectares)", y= "Frequency", title='Illinois Lakes')
```


<small> Figure 5. Distribution of Illinois lake size in hectares. </small>


In our analysis of the spatial distribution and size of lakes in Minnesota, Iowa, and Illinois, we found Minnesota exceeded the other two states both in the number of sites included in the LAGOS dataset and the size of lakes recorded in each state. There are 16446 sites in the LAGOS dataset for Illinois and Iowa combined. Meanwhile, there are 29038 Minnesota sites, exceeding the number of sites in Illinois/Iowa by 12572 sites. Furthermore, Minnesota lake size was generally greater than Iowa and Illinois lake size. Though, the distributions of lake size for all states are positively skewed. 


### Additional data sources {-}


Additional useful sources for assessing reservoir and lake size:


*The Global Lake area, Climate, and Population (GLCP) dataset is comprised of lake surface area data (from the datasets listed below), as well as temperature, precipitation, and population data (Meyer et al. 2020).

*The HydroLAKES dataset combines information from multiple lake datasets, including NASA SRTM, Water Body Data, and the Global Lakes and Wetlands Database (Meyer et al. 2020). This dataset consists of shapefiles with attributes such as lake surface area, total volume, average depth, geographic coordinates of pour points, and more (Meyer et al. 2020). 

*The Global Surface Water Dataset, derived from LANDSAT imagery and hosted by the Joint Research Centre (JRC), provides information regarding surface water area for lakes, as well as rivers, streams, and wetlands (Meyer et al. 2020). A subset of this data based upon yearly water classification history is available via Google Earth Engine (Meyer et al. 2020). 

*Also, the [Central Midwest Water Science Center](https://www.usgs.gov/centers/cm-water/science/real-time-data-links?qt-science_center_objects=0#qt-science_center_objects) is another resource, for Iowa and Illinois lake data specifically.


## References


Meyer, MF, Labou, SG, Cramer, AN, Brousil, MR, & Luff, BT. (2020). The global lake area, climate, and population dataset. Scientific Data, 7(1), 1–12. https://doi.org/10.1038/s41597-020-0517-4

