# Geospatial Analysis 


```{r, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(RApiSerialize)
library(ggthemes)
library(scales)
```


## Introduction


This project served as an introduction to spatial analysis in R and consisted of two parts. In part I, we utilized data from the LAGOS program to investigate the spatial distribution and size of lakes in midwestern US states. 

Project objectives: 


## Methods 


### Data Acquisition


```{r data-read}
#Lagos download script
# LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())

#Load in lagos data
lagos <- lagosne_load()

#Parse out lake centroid information
lake_centers <- lagos$locus
# load('lake_centers.Rdata')
```


### Data Processing


```{r include=F}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))
```


```{r}
#Create spatial object for lakes 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

#Subset number of lakes for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)
```


### Subset to only Minnesota


```{r include=F}
#Create states variable 
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)
```


## Results


```{r}
#Select only Minnesota from states data 
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plot the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha',at=c(0,100,250,500,1000,2500,5000,10000,125000), layer.name='Lake area (ha)')
```


## 1) Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r}
#Show a map outline of Iowa and Illinois 
i_states <- states %>% 
  filter(name %in% c('Iowa','Illinois')) %>% 
  st_transform(2163)

mapview(i_states, layer.name='States of Interest')
```



## 2) Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

```{r}
#Subset LAGOS data
istate_lakes <- spatial_lakes[i_states,]

nrow(minnesota_lakes)-nrow(istate_lakes)
```

There are 16446 sites in Illinois and Iowa combined. Meanwhile, there are 29038 Minnesota sites, exceeding the number of sites in Illinois/Iowa by 12572 sites.


## 3) What is the distribution of lake size in Iowa vs. Minnesota?


```{r}
# states_lagos <- lagos$state 

iowa <- states %>%
  filter(name == 'Iowa') %>%
  st_transform(2163)

#Subset lakes based upon spatial position
iowa_lakes <- spatial_lakes[iowa,]
  
#Create histogram of Iowa lake size 
ggplot(iowa_lakes, aes(x=lake_area_ha))+ 
  geom_histogram(bins=40, alpha=0.5)+ 
  theme_few()+
  scale_x_log10(labels = label_comma(accuracy = 1))+
  labs(x= "Lake size (hectares)", y= "Frequency", title='Iowa Lakes')

#Create histogram of Minnesota lake size
ggplot(minnesota_lakes, aes(x=lake_area_ha))+ 
  geom_histogram(bins=40, alpha=0.5)+ 
  scale_x_log10(labels = label_comma(accuracy = 1))+
  labs(x= "Lake size (hectares)", y= "Frequency", title='Minnesota Lakes')
```


The distributions of lake size for both Iowa and Minnesota are positively skewed. Generally, Iowa lake size is smaller than Minnesota.


<small> Figure . Interactive plot of Iowa and Illinois lakes by lake area in hectares. </small> 

```{r}
#Create interactive plot of Iowa and Illinois lakes 
#Color by lake area in hectares
istate_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) %>%
  mapview(., zcol = 'lake_area_ha', at=c(0,100,250,500,1000,2500,5000,10000), layer.name='Lake area (ha)', canvas=TRUE)
```



### Additional data sources: reservoirs and natural lakes size {-}


*The Global Lake area, Climate, and Population (GLCP) dataset is comprised of lake surface area data (from the datasets listed below), as well as temperature, precipitation, and population data (Meyer et al. 2020).

*The HydroLAKES dataset combines information from multiple lake datasets, including NASA SRTM, Water Body Data, and the Global Lakes and Wetlands Database (Meyer et al. 2020). This dataset consists of shapefiles with attributes such as lake surface area, total volume, average depth, geographic coordinates of pour points, and more (Meyer et al. 2020). 

*The Global Surface Water Dataset, derived from LANDSAT imagery and hosted by the Joint Research Centre (JRC), provides information regarding surface water area for lakes, as well as rivers, streams, and wetlands (Meyer et al. 2020). A subset of this data based upon yearly water classification history is available via Google Earth Engine (Meyer et al. 2020). 

*Also, the [Central Midwest Water Science Center](https://www.usgs.gov/centers/cm-water/science/real-time-data-links?qt-science_center_objects=0#qt-science_center_objects) is another resource, for Iowa and Illinois lake data specifically.


## References


Meyer, MF, Labou, SG, Cramer, AN, Brousil, MR, & Luff, BT. (2020). The global lake area, climate, and population dataset. Scientific Data, 7(1), 1–12. https://doi.org/10.1038/s41597-020-0517-4

