# Geospatial Analysis: Water Quality 


```{r, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(lubridate) #For dealing with date and time
library(ggthemes)
```


## Introduction






## Methods 


Similar to Geospatial Analysis part I, we retrieved lake water quality data using the LAGOSNE R package. 


### Data Acquisition {-}
```{r, cache=TRUE, include=F}
#Lagos download script
#lagosne_get(dest_folder = LAGOSNE:::lagos_path(),overwrite=T)

#Load in lagos
lagos <- lagosne_load()


#Grab the lake centroid info
lake_centers <- lagos$locus

# Make an sf object 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326)

#Grab the water quality data
nutr <- lagos$epi_nutr

#Look at column names
#names(nutr)
```


```{r}
#Select clarity variables from LAGOS water quality data 
clarity_only <- nutr %>%
  select(lagoslakeid,sampledate,chla,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))
```


### Data Processing {-}


```{r}
#Filter data to remove na values from variables of interest: Chla and Secchi Disk Depth
chla_secchi <- clarity_only %>%
  filter(!is.na(chla),
         !is.na(secchi))

#Include only lakes with at least 200 observations for Chla and Secchi Disk Depth
chla_secchi_200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)
```


```{r}
#Join water quality data to spatial data
spatial_200 <- inner_join(spatial_lakes,chla_secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')
```



```{r}
### Take the mean chl_a and secchi by lake
mean_values_200 <- chla_secchi_200 %>%
  # Take summary by lake id
  group_by(lagoslakeid) %>%
  # take mean chl_a per lake id
  summarize(mean_chl = mean(chla,na.rm=T),
            mean_secchi=mean(secchi,na.rm=T)) %>%
  #Get rid of NAs
  filter(!is.na(mean_chl),
         !is.na(mean_secchi)) %>%
  # Take the log base 10 of the mean_chl
  mutate(log10_mean_chl = log10(mean_chl))

#Join datasets
mean_spatial <- inner_join(spatial_lakes,mean_values_200,
                          by='lagoslakeid') 

#Make a map
mapview(mean_spatial,zcol='log10_mean_chl')
```


# Class work

## 1) What is the correlation between Secchi Disk Depth and Chlorophyll a for 
sites with at least 200 observations?

- Here, I just want a plot of chla vs secchi for all sites 

```{r}
#Plot secchi disk depth by cholorphyll a 
ggplot(data=chla_secchi_200, aes(x = secchi, y = chla)) +
  geom_point(shape = 1, color="#48680E")+ 
  theme_few()+
  labs(y=expression("Chlorophyll A"~(mu*"g/L")), x="Secchi Disk Depth (m)")+ 
  xlim(0,16)
```


```{r echo=TRUE, include=FALSE}
cor.test(chla_secchi_200$chla,chla_secchi_200$secchi)
```

There is a negative correlation between secchi disk depth and chlorophyll a levels, where shallower secchi disk depths are associated with higher chlorophyll a levels. 


## Why might this be the case? 

High cholorphyll a concentrations are associated with high density(s) of photosynthetic organisms, such as algae. Meanwhile, secchi disks are used to assess water turbidity, where the depth of disk disappearance indicates the transparency of the water (Fuller and Minnerick 2007). Where algae abundance is high, the depth where light can penetrate a water body diminishes; thus, secchi disk depth decreases (Fuller and Minnerick 2007).

Fuller, LM & Minnerick, RJ (2007). Predicting Water Quality by Relating Secchi-Disk Transparency and Chlorophyll a Measurements to Landsat Satellite Imagery for Michigan Inland Lakes, 2001–2006. USGS, August 2007, 1–4. [http://pubs.usgs.gov/fs/2007/3022/pdf/FS2007-3022.pdf]


## 2) What states have the most data? 

### 2a) First you will need to make a lagos spatial dataset that has the total 
number of counts per site.

```{r spatial object lagos, cache=TRUE}
#Create lagos spatial dataset with total number of counts per site 

site_counts <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n())

#Add geographic information to site counts data 

geo_counts <- inner_join(site_counts,lake_centers, by= 'lagoslakeid')%>% 
  select(lagoslakeid,nhd_long,nhd_lat, count, secchi, chla)

#Create spatial object for lagos dataset 

spatial_counts <- st_as_sf(geo_counts,coords=c('nhd_long','nhd_lat'),
                          crs=4326)
  
```


### 2b) Second, you will need to join this point dataset to the us_boundaries data.

```{r}
#Use a spatial join to combine point dataset and us_boundaries data

#Parse out state boundaries from us_boundaries data   

states <- us_states()

#Use spatial join to overlay point dataset (original lagos spatial dataset) and state boundaries 

overlay <- st_join(spatial_counts,states)

```


### 2c) Then you will want to group by state and sum all the observations in 
that state and arrange that data from most to least total observations per state. 

```{r, cache=TRUE}
# Group by state and sum number of observations by state

summed_overlay <- overlay %>%
  group_by(state_name) %>%
  summarize(sum_count = sum(count)) %>%
  arrange(desc(sum_count))

#Print first six rows of dataframe

knittr::kable(head(summed_overlay))

```


```{r echo=TRUE,include=FALSE}
#Plot/map number of observations per state

mapview(summed_overlay,zcol='sum_count', layer.name = 'Observations per state',canvas=TRUE)
```


The top five states with the most data include: Minnesota, Vermont, New York, Wisconsin, and Rhode Island. 


##3 Is there a spatial pattern in Secchi disk depth for lakes with at least 200 
observations?

```{r}
#Filter by 200 or more observations
#Calculate the mean secchi depth by lake

secchi200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200) %>% 
  summarize(mean_secchi = mean(secchi))
  
#Add geographic information, joining by lagoslakeid

secchi200_locus <- inner_join(secchi200, lake_centers, by='lagoslakeid') %>%
  select(lagoslakeid, nhd_long, nhd_lat, mean_secchi)

#Create a spatial object with lagoslakeid and mean secchi data

spatial_secchi200 <- st_as_sf(secchi200_locus,coords=c('nhd_long','nhd_lat'),
                          crs=4326)

#Visualize secchi data within interactive plot/map

mapview(spatial_secchi200, zcol='mean_secchi', layer.name='Secchi Disk Depth (m)', canvas=TRUE)

```


I do not find an exceptionally strong spatial pattern in secchi disk depth in lakes with at least 200 observations, given the number of observations above 200 is small. However, the patch of lakes near/surrounding the Minneapolis-Saint Paul sprawl in Minnesota have broadly shallower secchi disk depths. Meanwhile, lakes in more remote areas in northern Wisconsin and near the Adirondacks in New York have generally deeper secchi disk depths. Despite a weaker spatial pattern, it is probable lakes within more remote areas have generally deeper secchi depths and urban areas shallower depths. 





