---
title: "Introduction to Environmental Data Science: Final Project"
author: "Amanda Hastings"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
output: 
  bookdown::gitbook: default
# github-repo: amandahastings/ESS580_Final_Project
description: "This is book consists of six different projects completed in ESS580: Introduction to Environmental Data Science at Colorado State University in Spring 2022."
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(bookdown)
```


```{r,include=FALSE}
# knitr::write_bib(c(
#   .packages(), 'bookdown', 'knitr', 'rmarkdown'
# ), 'packages.bib')
```

# Preface {-}


This book contains six different projects completed in ESS580: Introduction to Environmental Data Science taught by Dr. Matt Ross and Dr. Nathan Mueller at Colorado State University. Each chapter details the purpose, data acquisition, analyses (where applicable), and results of an individual project completed in the 2022 spring semester. For each project, Dr. Matt Ross and/or Dr. Nathan Mueller wrote the preliminary code and outlined project objectives. I completed the remaining portions of each project and compiled this book. 

## References {-}

<small>

R packages used consistently throughout this book include: 

```{r, include=F}
citation(package="tidyverse")
citation(package="ggthemes")
citation(package="broom")
citation(package="lubridate")
citation(package="dygraphs")
citation(package="xts")
citation(package="mapview")
citation(package="sf")
citation(package="USAboundaries")
```

Appelhans et al. (2021). mapview-package: Interactive viewing of spatial objects in R. R package version 2.10.0. 
  https://github.com/r-spatial/mapview

Arnold, JB. (2021). ggthemes: Extra Themes, Scales and Geoms for 'ggplot2'. R package version
  4.2.4. https://CRAN.R-project.org/package=ggthemes
  
Grolemund, G. and Wickham, H. (2011). Dates and Times Made Easy with lubridate. Journal of
  Statistical Software, 40(3), 1-25. URL https://www.jstatsoft.org/v40/i03/.

Mullen, LA. and Bratt, B. (2018). "USAboundaries: Historical and Contemporary Boundaries of the
  United States of America," Journal of Open Source Software, 3(23), 314.
  https://doi.org/10.21105/joss.00314.
  
Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. The R Journal
  10(1), 439-446. https://doi.org/10.32614/RJ-2018-009
  
Robinson, D., Hayes, A., and Couch, S. (2022). broom: Convert Statistical Objects into Tidy
  Tibbles. R package version 0.7.11. https://CRAN.R-project.org/package=broom
  
Ryan, JA. and Ulrich, JM. (2020). xts: eXtensible Time Series. R package version 0.12.1.
  https://CRAN.R-project.org/package=xts
  
Vanderkam, D., Allaire, JJ., Owen, J., Gromer, D., and Thieurmel, B. (2018). dygraphs:
  Interface to 'Dygraphs' Interactive Time Series Charting Library. R package version 1.1.1.6.
  https://CRAN.R-project.org/package=dygraphs

Wickham et al. (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43),
  1686. https://doi.org/10.21105/joss.01686

</small>
