---
title: "Introduction to Environmental Data Science: Final Project"
author: "Amanda Hastings"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
output: 
  bookdown::gitbook: default
# github-repo: amandahastings/ESS580_Final_Project
description: "This is book consists of six different projects completed in ESS580: Introduction to Environmental Data Science at Colorado State University in Spring 2022."
---

```{r, include=F}
library(bookdown)
```


```{r,include=FALSE}
# knitr::write_bib(c(
#   .packages(), 'bookdown', 'knitr', 'rmarkdown'
# ), 'packages.bib')
```

# Preface {-}


This book contains six different projects completed in ESS580: Introduction to Environmental Data Science taught by Dr. Matt Ross and Dr. Nathan Mueller at Colorado State University. Each chapter details the purpose, data acquisition, analyses, and results of an individual project completed in the 2022 spring semester. For each project, Dr. Matt Ross and/or Dr. Nathan Mueller wrote the preliminary code and outlined project objectives. Amanda Hastings completed the remaining portions of each project and compiled this book. 




